Leaf traits drive differences in biomass partitioning among major plant functional types

### 2015-01-07 manuscript.Rnw
More edits and refs.

Globally, forests store ca. 360 Pg C in live biomass \citep{panetal2011}, or about 40 years of current athropogenic emissions \citep{friedlingsteinetal2014}. The partitioning of this biomass among plant components is a major determinant of vegetation productivity and the rate of carbon cycling \cite{ise2010}. Biomass stored in woody stems has a long residence time \citep{luyssaert2008},  thus representing a significant long-term carbon storage component. By contrast, leaf biomass turns over quickly, and is a major input to the soil carbon cycle where the majority of carbon is released back to the atmosphere. Recently, the residence time of carbon in vegetation has been identified as a major source of uncertainty in model projections of effects of climate change on the global carbon sink \citep{friend2014}. There is little consensus in these global vegetation models (GVMs) on how biomass partitioning should be modelled \citep{franklin2012,dekauwe2014}, but there is no shortage of underlying theory \citep{cannelldewar1994,franklin2012}. We stress that biomass partitioning, the actual distribution of biomass among components, is distinct from biomass allocation, the flux of carbon into plant components. Biomass partitioning is estimated from allocation by including the rates of turnover of the various plant components (see Methods S1).

It is difficult to decide which partitioning models are superior, since GVMs are usually only tested against productivity or carbon storage data, and both are the outcome of multiple processes. The vast majority of GVMs simplify parameterization by considering a few plant functional types (PFTs) \citep{harrisonetal2010}. It is also common to assume that parameters related to biomass partitioning varies between PFTs, however, parameterization of these differences have been based on little empirical evidence.  Based on previous compilations of forest biomass estimates, it is known that gymnosperms have more leaf biomass per unit above-ground biomass (the 'leaf mass fraction', LMF) than angiosperms \citep{oneilldeangelis1981,enquistniklas2002,reich2014}. However, it is not known whether this is due to the angiosperm/evergreen divergence, or to inherent differences in leaf life span. If the latter is true, we would expect evergreen angiosperms to have more leaf biomass than deciduous angiosperms.

We argue that the lack of a large global database on biomass partitioning at the individual plant level has hampered progress in this field. Here, we use a unique database to lay the empirical base against which models can be benchmarked. Our new database (BAAD, Falster) includes over 20,000 observations contributed by 175 studies of biomass and size metrics for individual plants, covering the entire size range including saplings and the largest trees in the world, three major woody PFTs, three biomes (tropical, temperate, and boreal forests), and nine vegetation types (Fig. S1). Another unique aspect is that the database contains leaf mass as well as leaf area estimates, whereas previous large compilations only considered leaf mass \citep{oneilldeangelis1981,enquistniklas2002,reich2014}. It is unclear whether PFTs with higher leaf biomass will also have higher total leaf area, because PFTs differ in the ratio of leaf mass to area (LMA), a measure of the leaf construction cost \citep{poorter2009}. Leaf area itself is relevant because light interception acts on leaf area, which drives photosynthesis. A few classic studies have assumed that gymnosperms indeed carry much more leaf area than angiosperms \citep{chabothicks1982, bond1989}. Finally it is also unknown whether potential differences between PFTs in biomass partitioning are a result of inherent differences in morphology, or whether climate directly affects partitioning, and the PFTs simply occupy different climate spaces.

@@

Next, we decomposed LMF into two components, $LMF = M_F/M_T = M_F/A_S * A_S/M_T$, where $M_F/A_S$ is the ratio of leaf biomass to stem basal area (see Methods S2), and $A_S/M_T$ the ratio of stem basal area to total above-ground biomass. We found that the first term showed identical PFT patterns as LMF (Fig 2d and e), $M_F/A_S$ was highly variable between PFT and proportional to LMA, so that $A_F/A_S$ is invariant between PFTs. This decomposition is also convenient because the first term is nearly independent of plant height (Tables 1 and 2), allowing a more straightforward analysis of climate effects (not having to correct for height every time). The second term ($A_S/MT$) was found to be nearly unaffected by PFT, but strongly height-dependent (Fig. S4). It thus absorbs variation related to height and above-ground plant geometry, and is consistent with previous findings that this ratio is fairly conserved between plant species \citep{chave2005}.

We analyzed climate effects in two ways: with quantitative variables (mean annual precipitation; MAP, mean annual temperature; MAT), and by biome (tropical / temperate/ boreal), which is a simple classification taking into account both MAT and MAP (Fig. S1). Regardless of how we analyzed the data, we found only weak climate effects. We did not find differences between biomes in $M_F/A_S$ (Fig. 3), but some small inconsistent differences in $A_F/A_S$ (corresponding to slightly lower LMA in tropical vs. temperate evergreen angiosperms). Biome consistently explained very little variation when added to a statistical model in addition to PFT and plant height ($R^2$ increased by only 0.01 - 0.06, see Table 2). Likewise, we found weak and inconsistent effects of climate variables (MAP and MAT) on $M_F/A_S$ and $A_F/A_S$ (Fig. 4). These variables explained little variation when added to a statistical model in addition to PFT and plant height ($R^2$ increased by 0.01 and 0.11, respectively, see Table 1). The only exception was LMA, where the $R^2$ increased from 0.47 to 0.61 when climate variables were added, consistent with previous reports that LMA shows higher plasticity than biomass partitioning \citep{poorter2012}.

@@

A recent study compared allocation routines in a number of leading ecosystem models \citep{dekauwe2014}, and recommended constraining allocation by observed biomass fractions instead of using constant allocation fractions. We argue that these models should be constrained by patterns with plant size and PFT presented in our study. As a first approximation, GVMS should aim to keep leaf area to stem cross-sectional area constant, and vary biomass allocation accordingly. The leaf mass per area, already a parameter in most GVMs, then gives the ratio of leaf biomass to stem biomass. An example of such a model is LPJ \citep{sitchetal2003}, where the ratio of leaf area to sapwood area is a constant parameter, but LMA differs between PFTs. Models that use constant allocation fractions may also be able to mimic the patterns between PFTs observed here, but only if leaf life span differs in such a way to maintain a proportionality between LMF and LMA. In any case, it is relatively straightforward to compare model simulations to broad patterns in LMF with PFT and plant height as presented in this paper. The data we use here is freely available (Falster), which should aid model testing.

>>>>

Globally, forests store ca. 360 Pg C in live biomass \citep{panetal2011}, or about 40 years of current athropogenic CO\textsubscript{2} emissions \citep{friedlingsteinetal2014}. The partitioning of this biomass among plant components is a major determinant of vegetation productivity and the rate of carbon cycling \cite{ise2010}. Biomass stored in woody stems has a long residence time \citep{luyssaert2008},  thus representing a significant long-term carbon storage component. By contrast, leaf biomass turns over quickly, and is a major input to the soil carbon cycle where the majority of carbon is released back to the atmosphere \cite{ryanlaw2005}. Recently, the residence time of carbon in vegetation has been identified as a major source of uncertainty in model projections of effects of climate change on the global carbon sink \citep{friend2014}. There is little consensus in these global vegetation models (GVMs) on how biomass partitioning should be modelled \citep{franklin2012,dekauwe2014}, but there is no shortage of underlying theory \citep{cannelldewar1994,franklin2012}. We stress that biomass partitioning, the actual distribution of biomass among components, is distinct from biomass allocation, the flux of carbon into plant components. Biomass partitioning is estimated from allocation by including the rates of turnover of the various plant components (see Methods S1).

It is difficult to decide which partitioning models are superior, since GVMs are usually only tested against productivity or carbon storage data, and both are the outcome of multiple processes. The vast majority of GVMs simplify parameterization by considering a few plant functional types (PFTs) \citep{harrisonetal2010}. It is also common to assume that parameters related to biomass partitioning vary between PFTs, however, estimatesof these parameters have been based on little empirical evidence.  Based on previous compilations of forest biomass estimates, it is known that gymnosperms have more leaf biomass per unit above-ground biomass (the 'leaf mass fraction', LMF) than angiosperms \citep{oneilldeangelis1981,enquistniklas2002,reich2014}. However, it is not known whether this is due to the angiosperm/evergreen divergence, or to inherent differences in leaf life span. If the latter is true, we would expect evergreen angiosperms to have more leaf biomass than deciduous angiosperms.

We argue that the lack of a large global database on biomass partitioning at the individual plant level has hampered progress in this field. Here, we use a unique database to lay the empirical base against which models can be benchmarked. Our new database (BAAD, Falster) includes over 21,000 observations contributed by 175 studies of biomass and size metrics for individual plants, covering the entire size range including saplings and the largest trees in the world, three major woody PFTs, three biomes (tropical, temperate, and boreal forests), and nine vegetation types (Fig. S1). Another unique aspect is that the database contains leaf mass as well as leaf area estimates, whereas previous large compilations only considered leaf mass \citep{oneilldeangelis1981,enquistniklas2002,reich2014}. It is unclear whether PFTs with higher leaf biomass will also have higher total leaf area, because PFTs differ in the ratio of leaf mass to area (LMA), a measure of the leaf construction cost \citep{poorter2009}. Leaf area itself is relevant because light interception acts on leaf area, which drives photosynthesis. A few classic studies have assumed that gymnosperms indeed carry much more leaf area than angiosperms \citep{chabothicks1982, bond1989}. Finally it is also unknown whether potential differences between PFTs in biomass partitioning are a result of inherent differences in morphology, or whether climate directly affects partitioning, and the PFTs simply occupy different climate spaces.

@@

Next, we decomposed LMF into two components, $LMF = M_F/M_T = M_F/A_S * A_S/M_T$, where $M_F/A_S$ is the ratio of leaf biomass to stem basal area (see Methods S2), and $A_S/M_T$ the ratio of stem basal area to total above-ground biomass. We found that the first term showed identical PFT patterns as LMF (Fig 2d and e), $M_F/A_S$ was highly variable between PFT and proportional to LMA, so that $A_F/A_S$ is invariant between PFTs. This decomposition is also convenient because the first term is nearly independent of plant height (Tables 1 and 2), allowing a more straightforward analysis of climate effects (avoiding the need to correct for height every time). The second term ($A_S/MT$) was found to be nearly unaffected by PFT, but strongly height-dependent (Fig. S4). It thus absorbs variation related to height and above-ground plant geometry, and is consistent with previous findings that this ratio is fairly conserved between plant species \citep{chave2005}.

We analyzed climate effects in two ways: with quantitative variables (mean annual precipitation; MAP, mean annual temperature; MAT), and by biome (tropical / temperate/ boreal), a simple classification taking into account both MAT and MAP (Fig. S1). Regardless of how we analyzed the data, we found only weak climate effects. We did not find differences between biomes in $M_F/A_S$ (Fig. 3), but some small inconsistent differences in $A_F/A_S$ (corresponding to slightly lower LMA in tropical vs. temperate evergreen angiosperms). Biome consistently explained very little variation when added to a statistical model in addition to PFT and plant height ($R^2$ increased by only 0.01 - 0.06, see Table 2). Likewise, we found weak and inconsistent effects of climate variables (MAP and MAT) on $M_F/A_S$ and $A_F/A_S$ (Fig. 4). These variables explained little variation when added to a statistical model in addition to PFT and plant height ($R^2$ increased by 0.01 and 0.11, respectively, see Table 1). The only exception was LMA, where the $R^2$ increased from 0.47 to 0.61 when climate variables were added, consistent with previous reports that LMA shows higher plasticity than biomass partitioning \citep{poorter2012}.

@@

A recent study compared allocation routines in a number of leading ecosystem models \citep{dekauwe2014}, and recommended constraining allocation by observed biomass fractions instead of using constant allocation fractions. We propose that these models are constrained by patterns with plant size and PFT presented in this study. As a first approximation, GVMS should aim to keep leaf area to stem cross-sectional area constant, and vary biomass allocation accordingly. The leaf mass per area, already a parameter in most GVMs, then gives the ratio of leaf biomass to stem biomass. An example of such a model is LPJ \citep{sitchetal2003}, where the ratio of leaf area to sapwood area is a constant parameter, but LMA differs between PFTs. Models that use constant allocation fractions may also be able to mimic the patterns between PFTs observed here, but only if leaf life span differs in such a way to maintain a proportionality between LMF and LMA. In any case, it is relatively straightforward to compare model simulations to broad patterns in LMF with PFT and plant height as presented here. The data we use here is freely available (Falster), which should aid model testing.
